### 2022-11-06 tex/kits.tex
[ writing ] various thesis additions
[ build ] get thesis into building state
[ writing ] lots of background, before feedback on later chapters

>>>>

\def\SimpleKits{../agda/processed-latex/SimpleKits.tex}

%Explain:
%
%\begin{itemize}
%  \item Specific uses of renaming/substitution in $\lambda$-calculus semantics.
%  \item General role of renaming/substitution in abstract algebra/syntax with
%    binding.
%\end{itemize}

A basic operation on any syntax with variables is \emph{substitution} --- the
replacement of variables in a term by terms with the same type as the variables.
In a sense, this is the defining operation of variables --- a variable is a
placeholder for a term, or equivalently in logic, a hypothesis is a placeholder
for an arbitrary proof.
In a type theory or logic, terms can bind variables, and we will typically have
operational semantics rules combining a term binding a variable with a term that
is to be substituted into the place of that variable, like the $\beta$-rule for
$\lambda$-calculus functions.

While substitution has this extra role in a lot of the syntaxes with binding we
care about, variable-binding also significantly complicates the substitution
operation.
Substitution acts on the free variables of a term, replacing them by terms, but
binders mean that some subterms have \emph{more} free variables than our
original term.
This causes different challenges for different representations of terms.
For example, with named variables and shadowing, na\"{i}vely defined
substitution could fall foul of variable capture.
In our approach, based on de Bruijn indices, the difficulty is that an index $i$
outside a binder of $n$ variables corresponds to an index $n + i$ inside the
binder.
Therefore, when substituting under a binder, we must first increment any free
variables contained in terms we are substituting in, which is a form of
\emph{renaming}.
Renaming replaces each free variable by another free variable, and is a special
case of substitution.
We must, however, define renaming before substitution, so as to avoid the
definition of substitution being circular.
Renaming avoids a similar circularity because when renaming goes under a binder,
we only have to increment each variable being renamed in, rather than each
variable \emph{in each term} being substituted in.

In this section, I formally implement simultaneous renaming and substitution for
the terms defined in the previous section.
Simultaneous substitution turns out to have a simple definition, which
generalises into other algorithms over terms with binders.
The section concludes with a unified implementation of renaming and
substitution, leaving further generalisation to the next section.

### 2022-11-06 tex/linmech.tex
[ writing ] lots of background, before feedback on later chapters

\subsection{yalla}
\begin{itemize}
  \item Sequent calculus
  \item Specialised to linear logic
\end{itemize}

>>>>

\subsection{Yalla}

\Citet{laurent18} presents a collection of linear logics in a uniform style, and
various proofs relating them.
The logics share varying amounts of definitions and theorems --- for example,
the main linear logic is parametrised on whether to include mix rules and
whether to restrict exchange to cyclic permutations, whereas systems like the
Lambek calculus (with no exchange) and polarised linear logic are defined
separately from scratch.

The style used in Yalla is to realise sequents as lists of formulas.
The active formula tends to be forced to be the first formula in such a list,
with an explicit exchange rule being used to move such formulas into place.
\Citet{laurent18} explains why explicit permutations are necessary in such a
system, and why using multisets instead of lists would conflate distinct proofs
when there are repeated formulas in a sequent.
The system I will present in the next chapter avoids permutations essentially by
being a refinemet of the system presented in the previous chapter, which has an
ordered context and specific references (de Bruijn indices) into it.

Despite making sure to distinguish distinct proofs, the Yalla library does not
define any equational theories of proofs, so does not prove any results relying
on these distinctions.
While the complication of defining an equational theory in the presence of an
exchange rule is probably largely inevitable, having the exchange rule
introduces redundancy such that many equivalent proofs are not equal as data
structures.
The mechanisations presented in \cref{sec:simple} all sought to reduce this kind
of redundancy for intuitionistic logic, so that the only non-trivial equations
in the equational theory are $\beta$- and $\eta$-rules (i.e.\ computationally
interesting rules).
I will restore this property of the representation in \cref{sec:semirings}.

\begin{itemize}
  \item Sequent calculus
  \item Specialised to linear logic
\end{itemize}

### 2022-11-06 tex/prelim.tex
[ writing ] content sections to full draft
[ writing ] lots of background, before feedback on later chapters

@@

@@

\begin{example}
  The following give rise to multicategories:
  \begin{itemize}
    \item Sets and multi-ary functions (functions of the form
      $A_1 \times \cdots \times A_n \to B$)
    \item $K$-vector spaces and multilinear maps
    \item For any simple type theory with sequents of the form
      $\Gamma \vdash A$, where $\Gamma$ is a list of types and $A$ is a type,
      types and open terms (up to $\beta\eta$-equivalence) form a multicategory.
  \end{itemize}
\end{example}

@@

\begin{definition}
  A multicategory $\C$ gives rise to its \emph{category of contexts} $\C^*$ as
  follows:
  \begin{itemize}
    \item $\obj(\C^*) \coloneqq \List(\obj\C)$
    \item $\C^*\plr{\sum_i \Psi_i, \Delta} \coloneqq \prod_i \C(\Psi_i; \Delta_i)$
      for any $\Psi : \List(\List(\obj\C))$
    \item $\id^* : \Gamma \to^* \Gamma \coloneqq \overline\id$
    \item $(\sigma : \Delta \to^* \Theta) \circ^* (\rho : \Gamma \to^* \Delta) :
      (\Gamma \to^* \Theta) \coloneqq \lambda i.~\sigma_i \circ \lambda j.~\rho_{ij}$
      \todo{Still wrong}
  \end{itemize}
\end{definition}

@@

\begin{definition}
  TODO: \emph{symmetric monoidal category}
\end{definition}

\begin{definition}
  TODO: \emph{commutative monoid in a symmetric monoidal category}
\end{definition}

Algebra

>>>>

In this section, I recall some standard categorical and algebraic notions that
are used in the rest of this thesis.
I present most of these notions in terms of \emph{multicategories}, whose main
purpose I see as giving universal properties to monoidal products --- for
example, vector spaces and multilinear maps form a multicategory, and in this
multicategory we can state the universal property of the tensor product of
vector spaces.
However, a reader already familiar with monoidal categories presented some other
way may wish to ignore the parts about multicategories and just skim the other
definitions to make sure they seem familiar.

@@

\section{Categories and multicategories}

@@

\begin{example}
  The following give rise to multicategories:
  \begin{itemize}
    \item Sets and multi-ary functions (functions of the form
      $A_1 \times \cdots \times A_n \to B$)
    \item $K$-vector spaces and multilinear maps, for any field $K$
    \item For any simple type theory with sequents of the form
      $\Gamma \vdash A$, where $\Gamma$ is a list of types and $A$ is a type,
      types and open terms (up to $\beta\eta$-equivalence) form a multicategory.
  \end{itemize}
\end{example}

@@

\begin{definition}
  A multicategory $\C$ gives rise to its \emph{category of contexts} $\C^*$ as
  follows:
  \begin{itemize}
    \item $\obj(\C^*) \coloneqq \List(\obj\C)$.
    \item $\C^*\plr{\sum_i \Psi_i, \Delta} \coloneqq \prod_i \C(\Psi_i; \Delta_i)$
      for any $\Psi : \List(\List(\obj\C))$.
    \item $\id^* : \Gamma \to^* \Gamma \coloneqq \overline\id$.
    \item $(\sigma : \Delta \to^* \Theta) \circ^* (\rho : \Gamma \to^* \Delta) :
      (\Gamma \to^* \Theta) \coloneqq \lambda i.~\sigma_i \circ \rho|_{\sigma,i}$,
      where $\rho|_{\sigma,i}$ is the restriction of $\rho$ to those objects
      in the domain of $\sigma_i$.
  \end{itemize}
\end{definition}

@@

\begin{definition}
  A \emph{symmetric multicategory} is a multicategory $\C$ together with, for
  each permutation $\sigma : \Gamma \tilde\to \Delta$, an action
  $\sigma(-) : \C(\Delta; A) \to \C(\Gamma; A)$ respecting identity and
  composition of permutations.
\end{definition}

\begin{remark}
  It is tempting to define symmetric multicategories by replacing the lists of
  objects in the definition of multicategories by finite multisets.
  However, doing so entirely forgets the order of inputs to an operation, for
  example forcing every morphism with a type of the form $A, A \to B$ to be
  commutative.
  There is a way to refine this idea to get a correct definition, at the cost of
  requiring more sophisticated higher categorical notions.
\end{remark}

\begin{definition}
  A \emph{symmetric monoidal category} is a representable symmetric
  multicategory, analogously to how I defined a monoidal category.
\end{definition}

\begin{definition}
  A \emph{commutative monoid} in a symmetric multicategory is a monoid in the
  underlying multicategory such that $\mathrm{swap}(\mu) = \mu$, where
  $\mathrm{swap} : M, M \tilde\to M, M$ is the permutation swapping two
  elements.
\end{definition}

\section{Algebra}

### 2022-11-06 tex/terms.tex
[ writing ] lots of background, before feedback on later chapters

\missingfigure{LTree}
\Ctx{}

We start with a data type \AgdaDatatype{\_⊢\_} of intrinsically simply typed
terms.
Beside base types, the only type former we have is the function type constructor
\AgdaInductiveConstructor{\_`→\_}.
Contexts (of type \AgdaRecord{Ctx}) are implemented as the free magma on types
(\AgdaDatatype{Ty}).
Context concatenation is \AgdaFunction{\_++ᶜ\_}, and \AgdaFunction{[\_]ᶜ}
embeds types into contexts.
Typed variables in a context are given by \AgdaRecord{\_∋\_}.
A variable in
\AgdaBound{$\Gamma$}\AgdaSpace{}\AgdaRecord{$\ni$}\AgdaSpace{}\AgdaBound{A}
is given by a path \AgdaField{idx} to a type in \AgdaBound{$\Gamma$}, together
with a proof \AgdaField{tyq} that this type is equal to \AgdaBound{A}.
Variables embed into terms via the \AgdaInductiveConstructor{var} constructor.

>>>>

\ExecuteMetaData[\LTreetex]{LTree}
\Ctx{}

Our first data structure involving contexts is that of intrinsically typed
variables.
A variable of type
\AgdaBound{$\Gamma$}\AgdaSpace{}\AgdaRecord{$\ni$}\AgdaSpace{}\AgdaBound{A}
is given by a path \AgdaField{idx} to a type in \AgdaBound{$\Gamma$}, together
with a proof \AgdaField{tyq} that this type is equal to \AgdaBound{A}.
Variables embed into terms via the \AgdaInductiveConstructor{var} constructor of
the family \AgdaDatatype{\_⊢\_} of intrinsically simply typed terms.
The only other syntactic forms we consider for now are the eliminator and
constructor of function types \AgdaInductiveConstructor{\_`→\_} ---
\AgdaInductiveConstructor{app} and \AgdaInductiveConstructor{lam}.
Application \AgdaInductiveConstructor{app} takes two subterms of the appropriate
types, while the subterm of $\lambda$-abstraction \AgdaInductiveConstructor{lam}
is in an extended context \GA{} --- \AgdaBound{$\Gamma$} concatenated with a
singleton context containing the type \AgdaBound{A}.
